Values disclosures and trust in science: A replication study

### 2022-07-25 paper/paper.Qmd
as emailed to KE, 2022-07-25

Challenges to the value-free ideal argue that at least some social and political values may, or even should, play a role in the evaluation of scientific hypotheses.  Keeping with the example, the question of whether chemical X causes cancer has much more significant social and political implications than the question of whether chemical X fluoresces under ultraviolet light.  So, in the interest of protecting human health, it would be appropriate to *scientifically* accept the hypothesis that chemical X is carcinogenic based on relatively provisional evidence [@CranorSocialBenefitsExpedited1995; @ElliottNonepistemicValuesMultiple2014; @FernandezPintoLegitimizingValuesRegulatory2019].

@@

This objection is an empirical prediction: if scientists disclose their values, they will be perceived as less trustworthy.  @ElliottValuesEnvironmentalResearch2017 conducted an online survey study to evaluate this prediction. The authors found tentative evidence that disclosing values may reduce the perceived trustworthiness of a scientist, and that this effect may be moderated by whether or not participants share the same values as the scientist and whether or not the scientist reports findings contrary to their stated values. However, these authors collected data from a somewhat small sample using Amazon Mechanical Turk, and adopted an analytical approach that significantly diluted their sample across different conditions. In this paper, we report the results of a replication of @ElliottValuesEnvironmentalResearch2017 study 1, using a larger sample and a more statistically efficient analytical approach. The major results found by @ElliottValuesEnvironmentalResearch2017 serve as the basis for our hypotheses, described below.

@@

We tested the hypothesis of a modest correlation between values and ideology in two ways. First, to test (H1a) whether political liberals are more likely to prioritize public health over economic growth compared to conservatives, we conducted a Spearman's rank order correlation. Results revealed a significant correlation in line with the hypothesis, Spearman's $\rho $ = -.47, *p* < .001. Political liberals were more likely than conservatives to value public health over economic growth. To test (H1b) that a majority of political conservatives priortize public health over economic growth we cross-tabulated the data. Results revealed that, contrary to the hypothesis, slightly more than half of the self-reported political conservatives in our sample reported valuing economic growth (51.7%) over public health (48.3%). See @fig-part-values.

@@

Next, we tested the hypothesis that (H4) if the participant and scientist share the same values, the scientist is perceived as more trustworthy than if the participant and scientist do not share the same values. For this and related analyses, we only included data from participants who were assigned to either of the Disclosure conditions and self-reported their own values, reducing the sample to 567. Following this, we created a new Shared Values variable as a composite of the participants' reported values and the scientist's values. For this analysis, an a priori directed acyclic graphic (DAG) indicated that a univariate regression of participants' METI rating onto Shared Values would produce a biased estimate, that adjustments were required for both the participant's and scientist's values, and that including demographic variables would not effect this estimate [@fig-shared-dag].  While the univariate model was significant, adj. $R^2$= .029, *F*(1, 565) = 18.0, *p* < .001 [@tbl-shared model 1], after adjusting for scientist values Shared Values did not emerge as a significant predictor of participants' perceptions of the trustworthiness of the scientist [@tbl-shared models 2-4], indicating that the univariate estimate was indeed biased.  Rather, only the scientist's disclosed values significantly predicted how trustworthy participants rated the scientist, such that a scientist who disclosed valuing public health was rated as more trustworthy than a scientist who disclosed valuing economic growth.  This scientist values effect was robust across alternative model specifications, with an estimated effect of 0.6 (95% CI 0.4-0.8) [@tbl-shared models 2-5], again consistent with the DAG.

@@

To test whether the findings regarding consumer risk sensitivity vary as a function of participants' values, we regressed participants' METI ratings of the scientist in the stimuli onto the Conclusions condition, Participants' Values variable, and the Conclusions by Participants' values interaction term. The full model was significant, adj. $R^2$ = .173, *F*(3, 840) = 59.95, *p* < .001 ([@tbl-eb model 2]). As with the earlier analysis, results showed a main effect of the Conclusions condition, such that participants rated the scientist who reported that BPA does not cause harm as less trustworthy than the scientist who reported that BPA causes harm, $\beta$ = -0.75, *t*(840) = -4.51, *p* < .001. However, this effect was qualified by a significant interaction with participants' values, $\beta$ = -0.41, *t*(840) = -2.16, *p* = .031. Participants who prioritized public health and read about a scientist who concluded that BPA causes harm rated the scientist as more trustworthy ($M_{sd} = 5.56_{1.09}$) than participants with the same values who read about a scientist who concluded that BPA does not cause harm ($M_{sd} = 4.40_{1.24}$) ([@fig-conclusion-part]).

@@

To test whether the findings above about a hypothesized transparency penalty may vary based on participants' values, we regressed participants' METI ratings of the scientist onto the Disclosure condition variable, Participants' Values variable, and the Disclosure by Participants' Values interaction term. The full model was not significant, adj. $R^2$< .001, *F*(3, 840) = 1.19, *p* = .31 [@tbl-ec model 2]. As with the earlier analysis, our results do not provide evidence for a transparency penalty to the perceived trustworthiness of a scientist, either in general or interacting with participants' own reported values, $\beta$ = -0.23, *t*(840) = -1.07, *p* = 0.286.

@@

Without adjustments, Shared Values appears to have a substantial interaction with Participant Values:  Shared Values appears to increase perceived trustworthiness for participants who value public health, while decreasing perceived trustworthiness for participants who value economic growth [@fig-shared-part].  However, as indicated by our analysis for a potential shared values effect, the estimated effects for both shared values and participant values are biased if the model is not adjusted for scientist values.  (@fig-shared-dag and @tbl-shared show how this bias occurs.  Shared Values is a collider between Participant Values and Scientist Values; hence, if a model specification includes Participant Values and Shared values but not Scientist Values, there is an open path between Participant Values and METI, resulting in a biased estimate for Participant Values.)  Because Shared Values is the logical biconditional of Participant Values and Scientist Values, and the Shared Values $\times$ Participant Values interaction term is their logical conjunction, including all four variables (Participant, Scientist, and Shared Values, along with the interaction term) in a regression model creates perfect collinearity.

![Data, group means, and 95% confidence intervals for H5d, interaction of shared values and participant values.  Panels correspond to participant values.](`r here(out_dir, '03_shared_part.png')`){#fig-shared-part}

>>>>

Challenges to the value-free ideal argue that at least some social and political values may, or even should, play a role in the evaluation of scientific hypotheses.  Keeping with the example, the question of whether chemical X causes cancer has much more significant social and political implications than the question of whether chemical X fluoresces under ultraviolet light.  So, in the interest of protecting human health, it would be appropriate to *scientifically* accept the hypothesis that chemical X is carcinogenic for regulatory purposes based on relatively weak or provisional evidence [@CranorSocialBenefitsExpedited1995; @ElliottNonepistemicValuesMultiple2014; @FernandezPintoLegitimizingValuesRegulatory2019].

@@

This objection is an empirical prediction: if scientists disclose their values, they will be perceived as less trustworthy.  @ElliottValuesEnvironmentalResearch2017 conducted an online survey study to evaluate this prediction. The authors found tentative evidence that disclosing values may reduce the perceived trustworthiness of a scientist, and that this effect may be moderated by whether or not participants share the same values as the scientist and whether or not the scientist reports findings contrary to their stated values. However, these authors collected data from a somewhat small sample using Amazon Mechanical Turk, and adopted an analytical approach that diluted their sample across different conditions. In this paper, we report the results of a replication of @ElliottValuesEnvironmentalResearch2017 study 1, using a larger sample and a more statistically efficient analytical approach. The major results found by @ElliottValuesEnvironmentalResearch2017 serve as the basis for our hypotheses, described below.

@@

We tested the hypothesis of a modest correlation between values and ideology in two ways. First, to test (H1a) whether political liberals are more likely to prioritize public health over economic growth compared to conservatives, we conducted a Spearman's rank order correlation. Results revealed a significant correlation in line with the hypothesis, Spearman's $\rho$ = -.47, *p* < .001. Political liberals were more likely than conservatives to value public health over economic growth. To test (H1b) that a majority of political conservatives priortize public health over economic growth we cross-tabulated the data. Results revealed that, contrary to the hypothesis, slightly more than half of the self-reported political conservatives in our sample reported valuing economic growth (51.7%) over public health (48.3%). See @fig-part-values.

@@

Next, we tested the hypothesis that (H4) if the participant and scientist share the same values, the scientist is perceived as more trustworthy than if the participant and scientist do not share the same values. For this and related analyses, we only included data from participants who were assigned to either of the Disclosure conditions and self-reported their own values, reducing the sample to 567. Following this, we created a new Shared Values variable as a composite of the participants' reported values and the scientist's values. For this analysis, an a priori directed acyclic graphic (DAG) indicated that a univariate regression of participants' METI rating onto Shared Values would produce a biased estimate, that adjustments were required for both the participant's and scientist's values, and that including demographic variables would not effect this estimate (@fig-shared-dag).  While the univariate model was significant, adj. $R^2$= .029, *F*(1, 565) = 18.0, *p* < .001 (@tbl-shared model 1), after adjusting for scientist values Shared Values did not emerge as a significant predictor of participants' perceptions of the trustworthiness of the scientist (@tbl-shared models 2-4), indicating that the univariate estimate was indeed biased.  Rather, only the scientist's disclosed values significantly predicted how trustworthy participants rated the scientist, such that a scientist who disclosed valuing public health was rated as more trustworthy than a scientist who disclosed valuing economic growth.  This scientist values effect was robust across alternative model specifications, with an estimated effect of 0.6 (95% CI 0.4-0.8; @tbl-shared models 2-5), again consistent with the DAG.

@@

To test whether the findings regarding consumer risk sensitivity vary as a function of participants' values, we regressed participants' METI ratings of the scientist in the stimuli onto the Conclusions condition, Participants' Values variable, and the Conclusions by Participants' values interaction term. The full model was significant, adj. $R^2$ = .173, *F*(3, 840) = 59.95, *p* < .001 (@tbl-eb model 2). As with the earlier analysis, results showed a main effect of the Conclusions condition, such that participants rated the scientist who reported that BPA does not cause harm as less trustworthy than the scientist who reported that BPA causes harm, $\beta$ = -0.75, *t*(840) = -4.51, *p* < .001. However, this effect was qualified by a significant interaction with participants' values, $\beta$ = -0.41, *t*(840) = -2.16, *p* = .031. Participants who prioritized public health and read about a scientist who concluded that BPA causes harm rated the scientist as more trustworthy ($M_{sd} = 5.56_{1.09}$) than participants with the same values who read about a scientist who concluded that BPA does not cause harm ($M_{sd} = 4.40_{1.24}$; @fig-conclusion-part).

@@

To test whether the findings above about a hypothesized transparency penalty may vary based on participants' values, we regressed participants' METI ratings of the scientist onto the Disclosure condition variable, Participants' Values variable, and the Disclosure by Participants' Values interaction term. The full model was not significant, adj. $R^2$< .001, *F*(3, 840) = 1.19, *p* = .31 (@tbl-ec model 2). As with the earlier analysis, our results do not provide evidence for a transparency penalty to the perceived trustworthiness of a scientist, either in general or interacting with participants' own reported values, $\beta$ = -0.23, *t*(840) = -1.07, *p* = 0.286.

@@

Without adjustments, Shared Values appears to have a substantial interaction with Participant Values:  Shared Values appears to increase perceived trustworthiness for participants who value public health, while decreasing perceived trustworthiness for participants who value economic growth (@fig-shared-part).  However, as indicated by our analysis for a potential shared values effect, the estimated effects for both shared values and participant values are biased if the model is not adjusted for scientist values.  (@fig-shared-dag and @tbl-shared show how this bias occurs.  Shared Values is a collider between Participant Values and Scientist Values; hence, if a model specification includes Participant Values and Shared values but not Scientist Values, there is an open path between Participant Values and METI, resulting in a biased estimate for Participant Values.)  Because Shared Values is the logical biconditional of Participant Values and Scientist Values, and the Shared Values $\times$ Participant Values interaction term is their logical conjunction, including all four variables (Participant, Scientist, and Shared Values, along with the interaction term) in a regression model creates perfect collinearity.

![Data, group means, and 95% confidence intervals for H5-shared, interaction of shared values and participant values.  Panels correspond to participant values.](`r here(out_dir, '03_shared_part.png')`){#fig-shared-part}
